Improving the portuguese hyphenation rules

### 2023-04-18 article/article.tex
updates from alines

\title{Improving the portuguese hyphenation rules}
\author{Leonardo Araújo}
\begin{document}
\maketitle

@@

\section{Grammar rules for hyphenation in Portuguese}
When there is no space left in line for a whole word, it might be split in two
using a hyphen as an indication of such procedure. In Portuguese, hyphenations
are allowed on syllables boundaries. According to the Grammar
\cite{cunha2016,bergstrom2011}, some rules might still apply:
\begin{enumerate}
    \item\label{rule-di-triphthong} diphthong or triphthong should not be split
	(e.g. \emph{m\underline{\emph{ui}}-to}, \emph{Pa-ra-g\underline{\emph{uai}}});
    \item\label{rule-unstressed} the sequences \emph{ia}, \emph{ie}, \emph{io}, \emph{oa}, \emph{ua},
	\emph{ue} and \emph{uo}, when in final unstressed position, should not be split 
	(e.g. \emph{gló-r\underline{\emph{ia}}}, \emph{vi-tó-r\underline{\emph{ia}}}, 
	\emph{cá-r\underline{\emph{ie}}}, \emph{es-pé-c\underline{\emph{ie}}}, 
	\emph{Má-r\underline{\emph{io}}}, \emph{má-g\underline{\emph{oa}}}, 
	\emph{ré-g\underline{\emph{ua}}}, \emph{tê-n\underline{\emph{ue}}}, 
	\emph{con-tí-g\underline{\emph{uo}}}, \emph{am-bí-g\underline{\emph{uo}}});
    \item\label{rule-c-clusters} consonant clusters starting a syllable should not be
	split (e.g. \emph{\underline{\emph{pn}}eu-má-ti-co}, \emph{\underline{\emph{ps}}i-có-lo-go}, 
	\emph{\underline{\emph{mn}}e-mô-ni-co});
    \item\label{rules-digraphs-ns} the digraphs \emph{ch}, \emph{lh}, \emph{nh}, \emph{gu}, \emph{qu} 
	should not be split (e.g. \emph{ra-\underline{\emph{ch}}ar},
	\emph{a-bro-\underline{\emph{lh}}os}, \emph{ma-\underline{\emph{nh}}ã},
	\emph{\underline{\emph{gu}}i-tar-ra}, \emph{\underline{\emph{qu}}in-to});
    \item\label{rule-nasalization} a vowel and its following nasalization marker (a graphic
	nasal consonant) should not be split (e.g. \emph{\underline{\emph{am}}-bição},
	\emph{m\underline{\emph{an}}-cha});
    \item\label{rule-rising} rising diphthong should not be split (e.g.
	\emph{a-b\underline{\emph{ai}}-xo}; \emph{c\underline{\emph{au}}-te-la}, 
	\emph{pa-p\underline{\emph{éi}}s}, \emph{cha-p\underline{\emph{éu}}}, 
	\emph{pre-f\underline{\emph{ei}}-to}, \emph{r\underline{\emph{eu}}-nir},
	\emph{n\underline{\emph{oi}}-te}; \emph{ca-la-b\underline{\emph{ou}}-ço}, 
	\emph{as-te-r\underline{\emph{ói}}-de}; 
	\emph{re-tri-b\underline{\emph{ui}}});

@@

>>>>

\title{Improving the portuguese hyphenation rules}
\author{Leonardo Araújo \and Aline de Lima Benevides}
\begin{document}
\maketitle

@@

\section{Grammar rules for hyphenation in Portuguese}

There is a wide range of writing systems, including logographic, syllabic, and
alphabetic writing systems. We might group the writing systems into three
cateories: alphabetic, phonemic and logographic. Although they are distinct
categories, some writing systems be built on an interplay of these categories
\cite{coulmas2003}. The principles that guide an alphabetic writing system or
a phonemic writing system can vary significantly based on the specific language
and its history, but we might highlight some of the key points: 
\begin{enumerate*}[label=\arabic*)]
    \item the most usual principle consists on representing the sounds of a
	language with written symbols, using one or a combination of symbols to
	represent each sound (e.g. the orthographic system of Portuguese uses
	an alphabet of 27 letters that aim to represent different sounds of the
	language); 
    \item etymology also plays an important role, where the
	spelling of a word reflects its origins and history (e.g. the French
	language often reflects the Latin roots of words in its spelling);
    \item morphology is used as a guide in structuring many languages, where
	they use specific letters or symbols to indicate word endings,
	prefixes, or suffixes (e.g. Russian uses different forms of the
	Cyrillic alphabet to indicate gender and case in its nouns); 
    \item the way a language has evolved over time also contribute (e.g. the 
	spelling of many English words has changed over time to reflect changes 
	in pronunciation);
\end{enumerate*}
The principles that guide orthographic systems for languages with logographic
writing systems like Chinese ideograms, cuneiform writing, and Egyptian
hieroglyphs differ significantly from those of alphabetic writing systems. In
logographic writing systems, individual symbols or characters represent entire
words or ideas rather than phonemes or sounds. As a result, the principles that
guide their orthographic systems are based more on semantic and visual
principles rather than phonemic principles.

As a result of the alphabetical writing system of the Portuguese, its spelling
is guided by phonological principles \cite{cagliari2015}. The fact that
spelling reflects, to some extent, the way words are pronounced, means that the
hyphenation of words is also influenced by the phonemic system, where words are
divided into syllables. It is noteworthy that different languages divide words
based on different principles, for example, English tends to be guided by
morphological principles, such as \emph{walk-ing}, \emph{un-happy},
\emph{work-s} and \emph{ear-ly}. Also, other factors affect word hyphenation in
English, such as the presence of doubled consonants and digraphs. There are also 
a myriad of exceptions that might be grasped into rules.
% https://english.stackexchange.com/questions/385/what-are-the-rules-for-splitting-words-at-the-end-of-a-line

Merely stating that an orthographic system is guided by phonological issues
does not necessarily mean that its hyphenation rules directly reflect the way a
word is pronounced. This is because the orthographic system operates according
to its own specific rules. For instance, although vowel clusters that produce a
single sound (i.e., digraphs) are pronounced within a single syllable in words
like \emph{chave}, \emph{ilha}, \emph{shampoo}, \emph{carro}, and \emph{bird},
their representation in writing determines how they are separated.
Specifically, different consonants within a digraph must remain together, while
different consonants are separated. As a result, we have \emph{i-lha} but \emph{car}.

%When there is no space left in line for a whole word, it might be split in two
%using a hyphen as an indication of such procedure. 
In Portuguese, hyphenations are allowed on syllables boundaries. According to
the Grammar \cite{cunha2016,bergstrom2011}, some rules might still apply:
\begin{enumerate}
    \item\label{rule-di-triphthong} diphthong or triphthong should not be split
	(e.g. \emph{m\underline{\emph{ui}}-to}, \emph{Pa-ra-g\underline{\emph{uai}}});
    \item\label{rule-unstressed} the sequences \emph{ia}, \emph{ie}, \emph{io}, \emph{oa}, \emph{ua},
	\emph{ue} and \emph{uo}, when in final unstressed position, should not be split 
	(e.g. \emph{gló-r\underline{\emph{ia}}}, \emph{vi-tó-r\underline{\emph{ia}}}, 
	\emph{cá-r\underline{\emph{ie}}}, \emph{es-pé-c\underline{\emph{ie}}}, 
	\emph{Má-r\underline{\emph{io}}}, \emph{má-g\underline{\emph{oa}}}, 
	\emph{ré-g\underline{\emph{ua}}}, \emph{tê-n\underline{\emph{ue}}}, 
	\emph{con-tí-g\underline{\emph{uo}}}, \emph{am-bí-g\underline{\emph{uo}}});
    \item\label{rule-c-clusters} consonant clusters starting a syllable should not be
	split (e.g. \emph{\underline{\emph{pn}}eu-má-ti-co}, \emph{\underline{\emph{ps}}i-có-lo-go}, 
	\emph{\underline{\emph{mn}}e-mô-ni-co});
    \item\label{rules-digraphs-ns} the digraphs \emph{ch}, \emph{lh}, \emph{nh}, \emph{gu}, \emph{qu} 
	should not be split (e.g. \emph{ra-\underline{\emph{ch}}ar},
	\emph{a-bro-\underline{\emph{lh}}os}, \emph{ma-\underline{\emph{nh}}ã},
	\emph{\underline{\emph{gu}}i-tar-ra}, \emph{\underline{\emph{qu}}in-to});
    \item\label{rule-nasalization} a vowel and its following nasalization marker (a graphic
	nasal consonant) should not be split (e.g. \emph{\underline{\emph{am}}-bição},
	\emph{m\underline{\emph{an}}-cha});
    \item\label{rule-rising} rising diphthong should not be split (e.g.
	\emph{a-b\underline{\emph{ai}}-xo}; \emph{c\underline{\emph{au}}-te-la}, 
	\emph{pa-p\underline{\emph{éi}}s}, \emph{cha-p\underline{\emph{éu}}}, 
	\emph{pre-f\underline{\emph{ei}}-to}, \emph{r\underline{\emph{eu}}-nir},
	\emph{n\underline{\emph{oi}}-te}; \emph{ca-la-b\underline{\emph{ou}}-ço}, 
	\emph{as-te-r\underline{\emph{ói}}-de}; 
	\emph{re-tri-b\underline{\emph{ui}}});

@@

\section{Updating the rules}

\section{Creating a new set of rules using Patgen}

\section{Comparing the set of rules}
